---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# praatpicture

<!-- badges: start -->
<!-- badges: end -->

If you've ever had to fiddle around with getting a Praat Picture to look and export just right in order to insert it in an R markdown file or slide deck, then this package is for you! It's a simple set of functions designed to emulate some common plots made in Praat Picture with base R plotting. It allows for flexibly combining waveforms, spectrograms, TextGrids, pitch tracks, formant tracks, and intensity tracks in a single plot. Waveforms and spectrograms are generated directly in R (the latter using the function `phonTools::spectrogram()`, which takes some time to plot and looks a little different from an equivalent Praat spectrogram, but still looks quite nice!). TextGrids, pitch tracks, formant tracks, and intensity tracks can be plotted as long as there are files with the `.TextGrid`, `.PitchTier`, `.Formant`, and `.IntensityTier` exported from Praat in the same directory and with the same base name as the sound file. These are loaded in using functions from the `rPraat` package. Default settings are as similar as possible to those in Praat, and the same with the different options for changing these settings. 

If you run into bugs or have suggested changes, please let me know!

## Installation

You can install the development version of `praatpicture` like so:

``` {r, eval=FALSE}
#install.packages('devtools')
devtools::install_github('rpuggaardrode/praatpicture')
```

## Examples

```{r}
library('praatpicture')
```

The Praat Picture I see (and use) most commonly simply consists of a relatively small waveform, a larger spectrogram, and an aligned TextGrid. Accordingly this is very simple to make with `praatpicture()`. Simply pass the name of a sound file with the `.wav` or `.mp3` extension to the function, and voila!

```{r}
praatpicture('inst/extdata/1.wav')
```

Often you won't want to plot an entire sound file, but only a small portion of it. This is no problem -- simply use the `start` and `end` arguments to tell R exactly what you want to plot. If we just want to plot the portion between 0.5--1.1 seconds of the above file, we do it like this:

```{r}
praatpicture('inst/extdata/1.wav', start=0.5, end=1.1)
```

By default, the first tier in the TextGrid is plotted. If we want to plot different tiers, we can tell R using the `tiers` argument. The `freqrange` argument controls the frequency limits of the spectrogram. We can use the `proportion` argument to control how much space is taken up by different frames. The `tfrom0` argument controls whether the x-axis starts from 0 or from the `start` time. 

Say we want to redo the plot with the 3rd and 4th tiers of the TextGrid, with frequencies up to 8000 Hz, showing the original time on the x axis, and with a bit less space for the waveform and more space for the TextGrid. It would look like this:

```{r}
praatpicture('inst/extdata/1.wav', start=0.5, end=1.1, tiers=c(3,4), 
             freqrange=c(0,8000), tfrom0=FALSE, proportion=c(20, 50, 30))
```

Other options for generating spectrograms, like in Praat, are `windowlength` (in seconds), `dynamicrange`, `timestep`, and `windowshape`.

If we want to reorder the components of the plot, or look at other components, we do so using the `frames` argument. If for example we wanted to plot a pitch track instead of a spectrogram, we do it like so:

```{r}
praatpicture('inst/extdata/1.wav', start=0.5, end=1.1, tiers=c(3,4), 
             frames=c('sound', 'pitch', 'TextGrid'))
```

I personally prefer the Praat's 'speckle' method of drawing pitch, which we can set using the `pitchtype` argument. We can also adjust the y-axis using `pitchrange`, and the scale using `pitchscale`. Say we want pitch between 100--200 Hz speckled in a log-scale, we do it like this:

```{r}
praatpicture('inst/extdata/1.wav', start=0.5, end=1.1, tiers=c(3,4), 
             frames=c('sound', 'pitch', 'TextGrid'), pitchtype='speckle', 
             pitchscale='logarithmic', pitchrange=c(100,200))
```

Additionally there's the argument `semitones_re` which sets the reference level if the scale is in semitones.

We could also plot formants instead of pitch:

```{r}
praatpicture('inst/extdata/1.wav', start=0.5, end=1.1, tiers=c(3,4), 
             frames=c('sound', 'formant', 'TextGrid'))
```
Or even just formants -- there's no requirement to plot waveform and TextGrid, we could just plot a single frame. If we wanted to draw formants with lines instead in a slightly narrower range (say, between 500--4000 Hz), we would use the `formantrange` and `formanttype` arguments.

```{r}
praatpicture('inst/extdata/1.wav', start=0.5, end=1.1, frames='formant', 
             formantrange=c(500,4000), formanttype='draw')
```

We could also have more frames -- for example, waveform, pitch track, intensity tier, and TextGrid at the same time, like this:

```{r}
praatpicture('inst/extdata/1.wav', start=0.5, end=1.1, tiers=c(3,4),
             frames=c('intensity', 'sound', 'pitch', 'TextGrid'),
             pitchtype='speckle', pitchrange=c(100,160))
```

Additionally, because `praatpicture`s are base R plots, it understands all global base plotting arguments. If, say, you want your Praat Picture to be red on a light grey background, using a larger monospaced font that's both bold and italic, there's literally nothing stopping you.

```{r}
praatpicture('inst/extdata/1.wav', start=0.5, end=1.1, tiers=c(3,4),
             frames=c('intensity', 'sound', 'pitch', 'TextGrid'), tier_names=F,
             pitchtype='speckle', pitchrange=c(100,160), bg='lightgrey', 
             bty='c', fg='red', font=4, ps=16, family='mono')
```

